---
title: "IMT 573: Lab 5 - Compare proportions"
author: "Madhusmita Oke"
output:
  html_document: default
---

# Collaborators
- Madhusmita

# Setup

Do whatever setup you do here, such as loading libraries

```{r setup, message=FALSE}
# Load standard libraries
library("tidyverse")
```

# 1 Simulations

## 1.1 to 1.4 Simulate for S0 and R repeatations, Check for CI 95%
H0: González has at 50% of total support. 

Set initial poll size S0 is 500, Polls to be repeated R times say 1000
```{r}
# Function to compute poll
poll <- function(poll_size) {
  s <- rbinom(poll_size, 1, 0.5)
  return(mean(s))
}

```

find mean of means for R=1000
```{r}
means <- sapply(1:1000, function(x) poll(500))
mean(means)

```


Computing 95% confidence level for H0
```{r}

c(quantile(means, 0.025), quantile(means, 0.975))
```

is 65% CI is inside or outside ? - It lies outside, So it indicates that poll size  was large, so we adjust the poll size
## 1.5 Repeat the above steps for S, R
```{r}
s1 <- 250
means2 <- sapply(1:1000, function(x) poll(s1))
#means
mean(means2)

c(quantile(means2, 0.025), quantile(means2, 0.975))
```

0.65 still lies outside for poll size 250, so we repeat the experiment.

```{r}
s2 <- 200
means3 <- sapply(1:1000, function(x) poll(s2))
#means
mean(means3)

c(quantile(means3, 0.025), quantile(means3, 0.975))
```


0.65 still lies outside for poll size 200, so we repeat the experiment.

```{r}
s3 <- 150
means4 <- sapply(1:1000, function(x) poll(s3))
#means
mean(means4)

c(quantile(means4, 0.025), quantile(means4, 0.975))
```

0.65 still lies outside for poll size 150, so we repeat the experiment.

```{r}
s4 <- 100
means5 <- sapply(1:1000, function(x) poll(s4))
#means
mean(means5)

c(quantile(means5, 0.025), quantile(means5, 0.975))
```

0.65 still lies outside for poll size 100, so we repeat the experiment.

```{r}
s5 <- 50
means6 <- sapply(1:1000, function(x) poll(s5))
#means
mean(means6)

c(quantile(means6, 0.025), quantile(means6, 0.975))
```


0.65 still lies outside for poll size 50, so we repeat the experiment.

```{r}
s6 <- 45
means7 <- sapply(1:1000, function(x) poll(s6))
#means
mean(means7)

c(quantile(means7, 0.025), quantile(means7, 0.975))
```


0.65 still lies outside for poll size 45, so we repeat the experiment.

```{r}
s7 <- 40
means8 <- sapply(1:1000, function(x) poll(s7))
#means
mean(means8)

c(quantile(means8, 0.025), quantile(means8, 0.975))
```


The result 65% for Harrell for poll size =40 is just at the boundary for CI 95% i.e. [0.35, 0.65]
Sample size is just right for S* = 40

## 1.6 Compute theoretical CI
```{r}
sample_size <- 40
p<-0.5
sigma2 = p*(1-p)/sample_size
sigma <- sqrt(sigma2)
sigma
```

Compute theoretical CI based on formula through central limit theorem on the properties of normal distribution 

```{r}

c(0.5 -  1.96 * sigma ,  0.5 + 1.96*sigma)
```

## 1.7 Based on your analysis, do you think Harrell is almost certain to be the winner? Do you see issues that your analysis did not address?

Based on analysis, the theoretical CI (95%) is almost equal to the computed CI for sample size 40. So we can say that Harrell is almost certain to win based on polls in audience of size 40. 


Our analysis is based of certain assumptions such as 65% winning chances for Harrell are representative of all Seattleites in question. Also, we are assuming that sufficient time has passed (we are closer to election voting deadline) when the 65% lead is calculated.Our analysis did not check for CI 99% in order to compute the likeliness of election result.
