---
title: "IMT 573: Lab 4 Create stochastic data"
author: "Madhusmita"
output:
  html_document: default
---

```{r setup, message=FALSE}
# Load standard libraries
library("tidyverse")
library("modeest")
```

# 1 Toss two coins

### 1.1. Which objects are you tossing? 
I tossed 1 coin twice , repeated it for 20 times.

```{r}
df_coin_toss <- tribble(
           ~TrialNo, ~Outcome, ~NoOfHeads,
1, 'TH', 1,
2, 'TT', 0,
3, 'TH', 1,
4, 'TT', 0,
5,  'HH',2,
6, 'TH', 1,
7, 'TH', 1,
8, 'HT',1,
9, 'TH',1,
10, 'TT',0,
11, 'TH',1,
12, 'HH',2,
13, 'HT',2,
14, 'HT',1,
15, 'HT',1,
16, 'TH',1,
17,'TT',0,
18, 'HH',2,
19, 'HH',2,
20, 'TH',1
)
```

### 1.2 Create a table about the number of heads in each trial

```{r}

df_coin_toss <- df_coin_toss[, !names(df_coin_toss) %in% "Outcome"]
df_coin_toss <- df_coin_toss %>% mutate(NoOfHeads = as.numeric(NoOfHeads))

df_coin_toss
```

### 1.3 show a summary table of your results (0 heads–how many times; 1 head–how many times..)

```{r}

zero_head <- count(df_coin_toss[df_coin_toss$NoOfHeads == 0, ])
#zero_head
one_head <- count(df_coin_toss[df_coin_toss$NoOfHeads == 1, ])
#one_head
two_head <- count(df_coin_toss[df_coin_toss$NoOfHeads == 2, ])
#two_head
df_summary_coin_toss <-
  data.frame(Heads = c(0, 1, 2),
             NoOfHeads = c(as.numeric(zero_head), as.numeric(one_head), as.numeric(two_head)))

df_summary_coin_toss
```


### 1.4 Compute mean of the results from coin toss experiment
```{r}
mean(df_coin_toss$NoOfHeads)
```
```{r}
modeest::mlv(df_coin_toss$NoOfHeads, na.rm=TRUE, method="mfv")
```

### 1.5 Compute variance of your results.

```{r}

# 1.compute mean
# 2. compute squares
# 3. compute sq of mean from 1.
# 4. compute mean of all squares from 2.
# 5. Use variance formula
df_mean <- mean(df_coin_toss$NoOfHeads)
df_sq <- df_coin_toss$NoOfHeads^2
df_sq_mean <- df_mean^2 # square of means
df_mean_sq <- mean(df_sq) # mean of squares 
variance <- df_mean_sq - df_sq_mean
variance
```

### 1.6 Plot Histogram of your results
```{r}
ggplot(df_coin_toss, aes(NoOfHeads)) + geom_histogram(fill="lightblue")
```


### 1.7 Can you explain why the histogram looks like it does? If you flip ordinary coins then you see the middle bar is (most likely) taller. Why?
If we flip ordinary coins (unbiased), the probability of getting only 1 head is 0.5 which is greater than probability of getting no heads (0.25) and both heads (0.25). Hence histogram (observations of head) would show a taller bar for 1 head scenario. This is demonstrated by the experiment where middle bar (1 head count) is taller (see above histogram).

### 1.8 Does it look symmetric, or right/left skewed?
The histogram looks symmetric 







