---
title: "IMT 573: Lab 2 - Make plots (and manipulate data)"
author: "Madhusmita Oke"
output:
  html_document: default
---

# Collaborators
- Madhusmita

# Setup

Do whatever setup you do here, such as loading libraries

```{r setup, message=FALSE}
# Load standard libraries
library("tidyverse")
library("dplyr")

```

# 1 Create a time variable


## 1.1 Load Data
```{r}

df_temperature <- read.csv("UAH-lower-troposphere-long.csv.bz2", sep = "\t")
nrow(df_temperature)
dim(df_temperature)
names(df_temperature)
head(df_temperature, 3)
```

## 1.2 Create Continuous time variable

```{r}
  # tail(df_temperature, 2)
  df_temperature <- df_temperature %>%
  select(year, month, region, temp) %>%
  mutate(timev = year + (month/12)) 
  
  dim(df_temperature)
  #head(df_temperature, 6)
  tail(df_temperature, 3)
```

# 2 Plot

```{r}

ggplot( df_temperature, aes(timev, temp, col=region)) +
geom_line()
```

## 2.1 Global Temperature over Time
```{r}
df_global_temp <- df_temperature[df_temperature$region == "globe",]
#dim(df_global_temp)
#length(unique(df_global_temp$year))

ggplot(df_global_temp, aes(timev, temp, col = region)) +
  geom_line(color = "blue")  
```


## 2.2 Plot North Hemisphere (nh) and Southern Hemisphere (sh) over time

```{r}
df_nh_sh <-
  df_temperature[df_temperature$region == "nh" |
                   df_temperature$region == "sh",]

ggplot(df_nh_sh, aes(timev, temp, col = region)) +
  geom_line()  
```


## 2.3 Plot only yearly global avg temperature

```{r}

df_yearly_avg_globe_temp <- df_temperature %>%
  select(temp, region, year) %>%
  filter(region == "globe") %>%
  group_by(year) %>%
  summarise(avg = mean(temp, na.rm = T))
#df_yearly_avg_globe_temp

ggplot(df_yearly_avg_globe_temp, aes(year, avg, col = avg, fill = avg)) +
  geom_col()  
```



## 2.4 Try other geom tools!

```{r}

ggplot(df_yearly_avg_globe_temp, aes(avg)) +
  geom_histogram(
    inherit.aes = T,
    stat = "bin",
    bins = 22,
    color = "red"
  )
```



## 2.5 Challenge
 Bar Plot with avg temperature over a decade - colour gradient to be used!
```{r}

df_decade_avg_temp <- df_temperature %>%
  select(temp, region, year) %>%
  mutate(decade = (year %/% 10)*10) %>%
  group_by(decade) %>%
  summarise(avg = mean(temp, na.rm = T))

#df_decade_avg_temp

  


ggplot(df_decade_avg_temp, aes(decade, avg, col = avg, fill = avg, scale_fill_gradient(low = "blue", high = "red"))) +
  geom_col()  
```

