---
title: 'IMT 573: PSet 8 - Predictive Modeling'
author: "Madhusmita Oke"
output: html_document
---

# Setup

Do whatever setup you do here, such as loading libraries

```{r setup, message=FALSE}
# Load standard libraries
library("tidyverse")
```

#  1 Is COMPAS fair?

## 1.1 Load the COMPAS data, and perform a basic sanity checks.

```{r}
df_compas_score_data <- read.delim("/Users/madhusmitaoke/MSIM-Academic/DataScience/IMT573-DSI/data/compas-score-data.csv.bz2")
names(df_compas_score_data)
dim(df_compas_score_data)
sapply(df_compas_score_data, range)
head(df_compas_score_data, 1)
```

There are 6172 observations with 8 columns / features : "age", "c_charge_degree", "race","age_cat", "sex", "priors_count", "decile_score",    "two_year_recid".

No missing values in data

## 1.2 Filter the data to keep only only Caucasians and African-Americans.

```{r}
unique(df_compas_score_data$race)
df_updated_compas_score_data <- df_compas_score_data %>% filter( race %in% c("African-American","Caucasian" ) )
unique(df_updated_compas_score_data$race)
```


Filtered data for to keep only Caucasians and African American

## 1.3 Create a new dummy variable based on the COMPAS’ risk score (decile_score), which indicates if an individual was classified as low risk (score 1-4) or high risk (score 5-10).
```{r}
df_updated_compas_score_data <- df_updated_compas_score_data %>% mutate(risk_score = ifelse(decile_score<=4, "low risk", "high risk"))
head(df_updated_compas_score_data, 2)
```

Created dummy variable "risk_score" where "low risk" indicates decile_score <=4 and "high risk" indicates decile_score>4.

<b>I am using df_updated_compas_score_data here onwards with updated data</b>


## 1.4 Now analyze the offenders across this new risk category:
### 1.4.a. What is the recidivism rate for low-risk and high-risk individuals?

<b>Calculating rate of recidivism for low risk as: (count of recidivists who are low risk / Total low risk)* 100</b>
```{r}
count_low_risk <- df_updated_compas_score_data %>%
  filter(risk_score == "low risk") %>%
  summarise(low_risk_count = n())

count_low_risk
recidivist_low_risk <- df_updated_compas_score_data %>%
  filter(two_year_recid == 1 & risk_score == "low risk") %>%
  summarise(recidivist_low_risk = n())

recidivist_low_risk

rate_low_risk <- (recidivist_low_risk/count_low_risk) * 100
rate_low_risk
```

Rate of recidivism for low risk is ~ 32%

<b> Calculating rate of recidivism for high risk as: (count of recidivists who are high risk / Total high risk)* 100 </b>

```{r}
count_high_risk <- df_updated_compas_score_data %>%
  filter(risk_score == "high risk") %>%
  summarise(high_risk_count = n())

count_high_risk
recidivist_high_risk <- df_updated_compas_score_data %>%
  filter(two_year_recid == 1 & risk_score == "high risk") %>%
  summarise(recidivist_high_risk = n())

recidivist_high_risk

rate_high_risk <- (recidivist_high_risk/count_high_risk) * 100
rate_high_risk
```
Rate of recidivism for high risk is ~ 63.4%

### 1.4.b What are the recidivism rates for African-Americans and Caucasians?
<b> Calculating rate of recidivism for African American as : (count of recidivists who are African American/ Total African American) *100 </b>

```{r}
count_af_am <- df_updated_compas_score_data %>%
  filter(race == "African-American") %>%
  summarise(af_am_count = n())

count_af_am
recidivist_af_am <- df_updated_compas_score_data %>%
  filter(two_year_recid == 1 & race == "African-American") %>%
  summarise(recidivist_af_am = n())

recidivist_af_am

rate_af_am <- (recidivist_af_am/count_af_am) * 100
rate_af_am
```

Rate of recidivism for African American is ~ 52.31 % 

<b>Calculating rate of recidivism for Caucasian as : (count of recidivists who are Caucasian/ Total Caucasian) *100  </b>
```{r}
count_casian <- df_updated_compas_score_data %>%
  filter(race == "Caucasian") %>%
  summarise(casian_count = n())

count_casian
recidivist_casian <- df_updated_compas_score_data %>%
  filter(two_year_recid == 1 & race == "Caucasian") %>%
  summarise(recidivist_casian = n())

recidivist_casian

rate_casian <- (recidivist_casian/count_casian) * 100
rate_casian
```

Rate of recidivism for Caucasian is ~39.09% 

## 1.5 Now create a confusion matrix comparing COMPAS predictions for recidivism (is/is not low risk) and the actual two-year recidivism and interpret the results. To keep things consistent, let’s call recidivists “positive”.

<b> NOTE : for ease of depicting confusion matrix, I've created variable "risk_tf" = 1 for "high risk" individuals and "risk_tf" = 0 for "low risk" individuals. This column risk_tf will act as "PREDICTED" and column two_year_recid will act as "ACTUAL" in the confusion matrix </b> 

recidivists -> positives -> "1"
non-recidivists -> negatives -> "0"

```{r}
df_updated_compas_score_data <- df_updated_compas_score_data %>% mutate(risk_tf = ifelse(risk_score =="high risk", 1, 0))
#df_updated_compas_score_data
conf_matrix <- table(Actual = df_updated_compas_score_data$two_year_recid, Predicted = df_updated_compas_score_data$risk_tf)

conf_matrix

TN <- conf_matrix[1]
TP <- conf_matrix[4]
FP <- conf_matrix[3]
FN <- conf_matrix[2]

#conf_matrix[1]
#conf_matrix[2]
#conf_matrix[3]
#conf_matrix[4]

```

Accuracy, Precision, Recall numbers calculation: 

```{r}

accuracy <- (TP + TN) / sum(conf_matrix)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)

cat("Accuracy :",accuracy, "\n")

cat("Precision:", precision, "\n")

cat("Recall:", recall, "\n")
```

Accuracy of 65.82% indicates that 65.82% of recidivists & non-recidivists where classified correctly by COMPAS.

Precision of 63.44% indicates that 63.44% of true positives ie. actual recidivists amongst all predicted recidvists - where predicted correctly by COMPAS. Basically 63.44% among all predicted recidivists were actually recidivists according to the model.

Recall of 64.52% indicates that 64.52% of true positives ie. actual recidivists amongst all actual recidivists (recidivists correcly predicted and wrongly predicted) in data - where predicted correctly by COMPAS. Basically, 64.52% of all actual recidivists where predicted recidivists by model.

     
## 1.6 Note the accuracy of the COMPAS classification, and also how its errors were distributed. Would you feel comfortable having a judge use COMPAS to inform your sentencing guidelines? At what point would the error/misclassification risk be acceptable for you?
The values of accuracy, precision and recall are close to each other. I would not be comfortable having a judget to use COMPAS to inform sentencing guidelines.

Percentage of actual recidivists amongst all predicted recidivists needs to improve a lot ,that is, Accuracy! It would be unfair if COMPAS keeps missing to classify "non-recidivists" as low risk and "recidivists" to high risk (biased on race as seen from the reading articles and analysis of results). But then the accuracy alone can be misguiding if the data used for model(maybe improved version) has class imbalance. We NEED BETTER recall as well - correctly classify true recidivists from all actual recidivists as crime rate may increase if recidivists keep getting away and not reprimanded appropriately.

I cant think of exact % acceptable recall or accuracy - maybe about > 95%.
It would be better for human judge to intervene if we know that algorithms are biased! Yes, we should be harsher with algorithmic results because many a times experts in the field along with their lived experience would provide a better sentence in judicial system.


## 1.7 Now, you repeat your confusion matrix calculation and analysis from 5. But this time do it separately for African-Americans and for Caucasians:

### 1.7.a. How accurate is the COMPAS classification for African-American individuals? For Caucasians?

<b> For African American </b>
```{r}
df_african_american <- df_updated_compas_score_data  %>%
  filter(race == "African-American")

#df_african_american

conf_matrix2 <- table(Actual = df_african_american$two_year_recid, Predicted = df_african_american$risk_tf)

conf_matrix2

TN <- conf_matrix2[1]
TP <- conf_matrix2[4]
FP <- conf_matrix2[3]
FN <- conf_matrix2[2]

accuracy <- (TP + TN) / sum(conf_matrix2)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)

cat("Accuracy :",accuracy, "\n")
cat("Precision :", precision, "\n")
cat("Recall :", recall, "\n")

```

We can see that accuracy for African American is low , 64.91% indicating misclassification of true recidivists and non-recidivists
Precision is low 64.95% - proportion of correctly predicted recidivists among all recidivists prediction
Recall is 71.52% - proportion of true predicted recidivists amongst actual recidivists.

<b> For Caucasian </b>
```{r}
df_caucasian <- df_updated_compas_score_data  %>%
  filter(race == "Caucasian")

#df_caucasian

conf_matrix3 <- table(Actual = df_caucasian$two_year_recid, Predicted = df_caucasian$risk_tf)

conf_matrix3

TN <- conf_matrix3[1]
TP <- conf_matrix3[4]
FP <- conf_matrix3[3]
FN <- conf_matrix3[2]

accuracy <- (TP + TN) / sum(conf_matrix3)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)

cat("Accuracy : ", accuracy, "\n")
cat("Precision:",  precision, "\n")
cat("Recall: ", recall, "\n")

  
  
```
For Caucasian as well, accuracy, precision and recall are on lower side.

### 1.7.b What are the false positive rates FPR = FP/N = FP/(FP + T N)?
<b> For African American </b>
```{r}
FP <- conf_matrix2[3]
TN <- conf_matrix2[1]
FPR <- FP/(FP + TN)
cat("False positive rate:", FPR)
```

<b> For Caucasian </b>
```{r}
FP <- conf_matrix3[3]
TN <- conf_matrix3[1]
FPR <- FP/(FP + TN)
cat("False positive rate:",FPR)
```


### 1.7.c. The false negative rates FNR = FN/P = FN/(FN + T P)?
<b> For African American </b>
```{r}
FN <- conf_matrix2[2]
TP <- conf_matrix2[4]
FNR <- FN/(FN + TP)
cat("False negative rate:", FNR)
```

<b> For Caucasian </b>
```{r}
FN <- conf_matrix3[2]
TP <- conf_matrix3[4]
FNR <- FN/(FN + TP)
cat("False negative rate:", FNR)
```


## 1.8  If you have done this correctly, you will find that COMPAS’s true negative and true positive percentages are similar for African-American and Caucasian individuals, but that false positive rates and false negative rates are different. Look again at the overall recidivism rates in the dataset for Black and White individuals. In your opinion, is the COMPAS algorithm ’fair’? Justify your answer.

We can see that accuracy for African American and Caucasian are similar but the false positive rate and false negative rate are different! In my opinion, COMPAS algorithm is not "fair". 
False positive rate would show % of individuals falsely predicted as recidivists : this rate is higher for African American as compared to Caucasian (42% vs 22% ).
False negative rate would show % of individuals falsely predicted as non-recidivists : this rate is higher for Caucasian as compared to African American (49% vs 28% )

These rates clearly show that African Americans are flagged as recidivists more than Caucasian. I feel although there are chances that people with either of the races may end up in recidivism but African American folks are flagged as high risk (predicted recidivists) more than the other race.

From the reading/ article mentioned in the PSet as well, I believe there has been a bias based upon race while making these COMPAS predictions (which is not even transparent about showing criteria saying it is proprietary characteristic )
Hence, I think COMPAS algorithm is not 'fair'.