---
title: 'IMT 573: Lab 3 - Get game scores from the web'
author: "Madhusmita Oke"
output: html_document
---


# Collaborators
Madhusmita Oke

# Setup

Do whatever setup you do here, such as loading libraries

```{r setup , message=FALSE}
# Load standard libraries
library("tidyverse")
library(rvest)

```

# 1 Scrape recent college football scores
## 1.1 Load and parse football page
```{r}
sports_url  <- "https://www.sports-reference.com/cfb/"
sports <- read_html(sports_url)
```

## 1.2 Find which element contains text "Recent Games" 
 On Inspecting, "Recent Games" is part of h2 element 
 
 
## 1.3 Find the class of the Div in which the above h2 element is contained
 The Div tag has class "game_summaries" which contains the h2 tag of "Recent Games"
 
## 1.4 Extract the DIV class  
```{r}
  summaries <- sports %>% html_element("div.game_summaries")
  summaries
```


## 1.5, 1.6 Which Games scores are embedded into which tags?
The scores such as Auburn 38 , Arkansas (17) 23 are embedded as part of a or anchor tag which is url to redirect to relevant web page


## 1.7 Extract the elements and print text!
```{r}
games_text <- summaries %>% html_elements("table.teams") %>% html_text()

games_text <- str_replace_all(games_text, "\\t", "") %>% str_replace_all(.,"\\n", "" ) 
games_text
```

## 1.8 Load values in dataframe using html_table
```{r}
table_content <- summaries %>% html_elements("table.teams") 
games_df <- data.frame()
for( tab in table_content){
  games_df <- rbind(games_df, tab %>% html_table())
}

dim(games_df)
head(games_df, 4)
```

## 1.9 Challenge

```{r eval=FALSE}
games_df <- games_df[, -which(names(games_df) == "X3")]

#games_df

score_card_df <- data.frame(Team1 = character(0), Score1 = numeric(0), Team2 = character(0), Score2 = numeric(0))

num_rows <- nrow(games_df)

for (i in 1:num_rows) {
  row_data <- games_df[i, ]  
 
  if(i%%2 == 1){
        score_card_df <- score_card_df %>%
          mutate(score_card_df$Team1 <- row_data$X1)
         score_card_df$Score1 <- row_data$X2)
        #score_card_df$Team1 <- row_data$X1
        #score_card_df$Score1 <- row_data$X2
  }else{
   
        print("")
  }
  
}


#score_card_df
```


